Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-08-20 ms.tex
removing excess line profile definition, changing mentions of "photon source frequency" to just x_s

 \begin{figure}
    \centering
    \includegraphics{xinit_threepanel_tau1e6.pdf}
    \caption{Again shifting the initial photon frequency, but at a lower optical depth of $\tau_0 = 10^6$. The shift in $x$ is a much larger fraction of the distance to the spectral peak $(a\tau_0)^{1/3}$, and thus the bias in the spectrum is much more drastic. The optical depth at the source frequency is $\tau_s = 10^6$, $7.7$, and $1.9$ for $x_s=0.0, 6.0,$ and $12.0$, respectively. } 
    \label{fig:sol_mc_xinit_lowtau}
\end{figure}

Since the solutions have been generalized to allow for a monochromatic source of photons at a frequency $\sigma_s$ away from line center, their agreement with Monte Carlo as a function of source frequency can also be characterized. Photons initialized closer to the line wing may travel larger mean free paths as the opacity is sensitively dependent on the line profile. The larger spatial diffusion contributes to greater escape probability, biasing the frequency distribution of escaping photons. In the limit that $|\rm x_s|$ becomes large, the distribution becomes a delta function in frequency as all photons escape the gas without scattering. Fig. \ref{fig:sol_mc_xinit} shows calculations performed for three initial photon frequencies $\rm x_s = 0.0, 6.0$, and $12.0$ for a line center optical depth of $\tau_0=10^7$. The bias in frequency distribution is slight for $\rm x_s = 6.0$ where the line profile is still relatively large and emitted photons remain trapped in the core, but the increase in spatial diffusion is much more substantial when photons are initialized with $\rm x_s=12.0$, further away from the line core. It is seen here that the discrepancy between the Monte Carlo data and $H_0$ (effectively the scale of the corrective $H_{\rm bc}$ term) becomes larger as $\rm x_s$ increases. Thus, for sufficiently large optical depths and emission away from line center, the importance of enforcing frequency-dependent boundary conditions at the surface of the sphere grows. The bias in the escaping spectrum is much more noticeable at lower optical depths, shown in Fig. \ref{fig:sol_mc_xinit_lowtau}, where a $\Delta x_s$ of 6 is a significantly larger fraction of the spectral peak location $(a\tau_0)^{1/3}$. Here it is more clear that as the photon initialization frequency becomes further from the optically thick core, the spectrum approaches a delta function in frequency at $\sigma_s$. Additionally, since the material is optically thin here, we expect there to be a stronger disagreement between the Monte Carlo and the analytic solution as derivations assumed large optical depths.

\begin{figure}
    \centering
    \includegraphics{xinit.pdf}
    \caption{$H_{\rm bc}$ solutions shown at three line center optical depths, $\tau_0=10^5, 10^7,$ and $10^9$, for photon source frequencies $\sigma_s=0.0, \tau_0, 2\tau_0$. Thicker lines indicate larger optical depth, and darker lines indicate a photon source frequency further from line center. The axes are scaled to show that the size of the corrective factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.}
    \label{fig:xinit}
\end{figure}

Fig. \ref{fig:xinit} shows how the boundary term scales with both optical depth and photon source frequency. The photon initialization frequency (in units of $\sigma$, as in the differential equation) is shifted in integer multiples of $\tau_0$. For clarity, only the positive side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, the size of the correction should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{1/3}$ scaling. The remaining discrepancy present is due to the width of the Doppler core with respect to the peak of the spectrum; at lower depths, this causes the assumptions about the functional form of the line profile to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{1/3}$, where $a$ is the ratio of natural to Doppler broadening and $\tau_0$ is the optical depth at line center. An additional amount of error of order $(a\tau_0)^{1/3}$ is introduced as the photon frequency $\sigma_s$ from line center is increased by a factor of $\tau_0$.

@@

The form of a single eigenmode $J_{nm}(\sigma)$ is oscillatory out to some turning point, $\sigma_{tp}$, at which point the function becomes evanescent. The location of the turning point can be found by ignoring the delta-function discontinuity at the source frequency $\sigma_s$ in Eq. \ref{eq:diffusion_plugged_in} and re-examining the resulting homogeneous differential equation. We obtain
\be
\frac{d^2J}{d\sigma^2} & = & \left[ \left( \frac{\kappa_n \Delta }{k} \right)^2 - \frac{3\phi \gamma\Delta^2}{ck}\right] J
\ee
where the line profile can be written approximately as $\phi = (2a/\pi)^{1/3}/(3 \Delta |\sigma|^{2/3})$. When the coefficient on the right hand side is positive, exponential growth or decay solutions are found (evanescence). This occurs on the line wings. When the coefficient on the right hand side is negative, oscillatory solutions are found (propagation), which occurs in the line core. The boundary between propagation and evanescence occurs at the turning point, given by
\be
\sigma_{tp} & = & \sqrt{\frac{2a}{\pi}}\left( \frac{k \gamma}{ \kappa_n^2 c \Delta} \right)^{3/2}
\ee
Thus, to ensure accuracy in the final sum of Eq. \ref{eq:Jrsigmat}, the bounds of $\sigma$ must be set sufficiently far outside of $\sigma_{tp}$ such that the function is small at the edges. The scale of an $e$-folding in $J_{nm}(\sigma)$ is $k/(\kappa_n \Delta) = \tau_0 / (\sqrt{\pi} n)$, so a grid of $\sigma$ is chosen that spans a large enough number of $e$-foldings that no oscillatory behavior is present at the boundaries of the calculation.

@@

The solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere originally derived by \cite{2006ApJ...649...14D} has been generalized to allow a monochromatic source of photons with frequencies away from line center. A correction to the $J=0$ surface boundary condition which allows the proper condition $J=\sqrt{3}H$ to be satisfied comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$, which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$ where $a$ is the damping parameter and $\tau_0$ is the line-center optical depth. Thus, for very large optical depths, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower optical depths. Additionally, since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Since the peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$, calculations performed at lower optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core. By comparison with Monte Carlo simulations, it is seen that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency is altered, provided a sufficiently large optical depth.

The time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency, the flux at the surface of the sphere is found as a function of time and photon frequency, expressed as a sum over many eigenmodes $n$ and $m$---spatial and frequency modes, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, biasing the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. When calculated as a cumulative distribution function using a rejection method, the fit can be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method. As will be shown in future work, these calculations can be implemented as an acceleration scheme for Monte Carlo resonance-line radiation transfer at optical depths that would nominally be computationally prohibitive.

>>>>

 \begin{figure}
    \centering
    \includegraphics{xinit_threepanel_tau1e6.pdf}
    \caption{Again changing $x_s$, but at a lower optical depth of $\tau_0 = 10^6$. The shift in $x$ is a much larger fraction of the distance to the spectral peak $(a\tau_0)^{1/3}$, and thus the bias in the spectrum is much more drastic. The optical depth at the source frequency is $\tau_s = 10^6$, $7.7$, and $1.9$ for $x_s=0.0, 6.0,$ and $12.0$, respectively. } 
    \label{fig:sol_mc_xinit_lowtau}
\end{figure}

Since the solutions have been generalized to allow for a monochromatic source of photons at a frequency $\sigma_s$ away from line center, their agreement with Monte Carlo as a function of source frequency can also be characterized. Photons initialized closer to the line wing may travel larger mean free paths as the opacity is sensitively dependent on the line profile. The larger spatial diffusion contributes to greater escape probability, biasing the frequency distribution of escaping photons. In the limit that $|\rm x_s|$ becomes large, the distribution becomes a delta function in frequency as all photons escape the gas without scattering. Fig. \ref{fig:sol_mc_xinit} shows calculations performed for three initial photon frequencies $\rm x_s = 0.0, 6.0$, and $12.0$ for a line center optical depth of $\tau_0=10^7$. The bias in frequency distribution is slight for $\rm x_s = 6.0$ where the line profile is still relatively large and emitted photons remain trapped in the core, but the increase in spatial diffusion is much more substantial when photons are initialized with $\rm x_s=12.0$, further away from the line core. It is seen here that the discrepancy between the Monte Carlo data and $H_0$ (effectively the scale of the corrective $H_{\rm bc}$ term) becomes larger as $\rm x_s$ increases. Thus, for sufficiently large optical depths and emission away from line center, the importance of enforcing frequency-dependent boundary conditions at the surface of the sphere grows. The bias in the escaping spectrum is much more noticeable at lower optical depths, shown in Fig. \ref{fig:sol_mc_xinit_lowtau}, where a $\Delta x_s$ of 6 is a significantly larger fraction of the spectral peak location $(a\tau_0)^{1/3}$. Here it is more clear that as $x_s$ becomes further from the optically thick core, the spectrum approaches a delta function in frequency at $\sigma_s$. Additionally, since the material is optically thin here, we expect there to be a stronger disagreement between the Monte Carlo and the analytic solution as derivations assumed large optical depths.

\begin{figure}
    \centering
    \includegraphics{xinit.pdf}
    \caption{$H_{\rm bc}$ solutions shown at three line center optical depths, $\tau_0=10^5, 10^7,$ and $10^9$, for photon source frequencies $\sigma_s=0.0, \tau_0, 2\tau_0$. Thicker lines indicate larger optical depth, and darker lines indicate $\sigma_s$ further from line center. The axes are scaled to show that the size of the corrective factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.}
    \label{fig:xinit}
\end{figure}

Fig. \ref{fig:xinit} shows how the boundary term scales with both $\tau_0$ and $\sigma_s$. The photon initialization frequency (in units of $\sigma$, as in the differential equation) is shifted in integer multiples of $\tau_0$. For clarity, only the positive side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, the size of the correction should become smaller with larger optical depths, following a $(a\tau_0)^{-1/3}$ scaling. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{1/3}$ scaling. The remaining discrepancy present is due to the width of the Doppler core with respect to the peak of the spectrum; at lower depths, this causes the assumptions about the functional form of the line profile to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{1/3}$, where $a$ is the ratio of natural to Doppler broadening and $\tau_0$ is the optical depth at line center. An additional amount of error of order $(a\tau_0)^{1/3}$ is introduced as the photon frequency $\sigma_s$ from line center is increased by a factor of $\tau_0$.

@@

The form of a single eigenmode $J_{nm}(\sigma)$ is oscillatory out to some turning point, $\sigma_{tp}$, at which point the function becomes evanescent. The location of the turning point can be found by ignoring the delta-function discontinuity at the source frequency $\sigma_s$ in Eq. \ref{eq:diffusion_plugged_in} and re-examining the resulting homogeneous differential equation. We obtain
\be
\frac{d^2J}{d\sigma^2} & = & \left[ \left( \frac{\kappa_n \Delta }{k} \right)^2 - \frac{3\phi \gamma\Delta^2}{ck}\right] J
\ee
where the line profile is approximated as in Eq. \ref{app:rteqn_derivation}\ref{eq:line_profile_wing}. When the coefficient on the right hand side is positive, exponential growth or decay solutions are found (evanescence). This occurs on the line wings. When the coefficient on the right hand side is negative, oscillatory solutions are found (propagation), which occurs in the line core. The boundary between propagation and evanescence occurs at the turning point, given by
\be
\sigma_{tp} & = & \sqrt{\frac{2a}{\pi}}\left( \frac{k \gamma}{ \kappa_n^2 c \Delta} \right)^{3/2}
\ee
Thus, to ensure accuracy in the final sum of Eq. \ref{eq:Jrsigmat}, the bounds of $\sigma$ must be set sufficiently far outside of $\sigma_{tp}$ such that the function is small at the edges. The scale of an $e$-folding in $J_{nm}(\sigma)$ is $k/(\kappa_n \Delta) = \tau_0 / (\sqrt{\pi} n)$, so a grid of $\sigma$ is chosen that spans a large enough number of $e$-foldings that no oscillatory behavior is present at the boundaries of the calculation.

@@

The solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere originally derived by \cite{2006ApJ...649...14D} has been generalized to allow a monochromatic source of photons with frequencies away from line center. A correction to the $J=0$ surface boundary condition which allows the proper condition $J=\sqrt{3}H$ to be satisfied comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$, which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$ where $a$ is the damping parameter and $\tau_0$ is the line-center optical depth. Thus, for very large optical depths, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower optical depths. Additionally, since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. Since the peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$, calculations performed at lower optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core. By comparison with Monte Carlo simulations, it is seen that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency $x_s$ is altered, provided a sufficiently large optical depth $\tau_0$.

The time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over many eigenmodes $n$ and $m$---spatial and frequency modes, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, biasing the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. When calculated as a cumulative distribution function using a rejection method, the fit can be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method. As will be shown in future work, these calculations can be implemented as an acceleration scheme for Monte Carlo resonance-line radiation transfer at optical depths that would nominally be computationally prohibitive.
